Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-19 paper.tex
edits to sec 2

\newcommand{\cOOcast}[3]{#1 \overset{#2}{\Rightarrow} #3}

@@

@@

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
	Designing a space-efficient cast representation that is good for both 
	mechanized metatheory and implementation is challenging. Existing solution 
	solutions have various strengths and weaknesses. We presents a new cast 
	representation, named hypercoercions, that is good for both. On our way to 
	prove the correctness of hypercoercions, we also step toward a framework 
	for proving correctness of cast representations.
\end{abstract}

@@

Around 2006, several groups of researchers proposed ways to integrate
dynamic typing and static typing, notably gradual typing
\citep{siek2006gradual}, hybrid typing \citep{flanagan2006hybrid},
migratory typing \citep{Tobin-Hochstadt:2006fk}, and multi-language
iteroperability \citep{Gray:2005ij,Matthews:2007zr}. Researchers
usually define the semantics of gradually typed languages by
translation to an intermediate language with casts, such as the blame
calculus \citep{wadler2009well} and other cast calculi
\citep{siek2009exploring}. Unfortunately, straightforward
implementations of casts on higher-order values (functions, objects,
etc.) impose significant runtime overheads that can change the
asymptotic time and space complexity of a program
\citep{herman2010space}. There are several known space-efficient cast
representations, with various strengths and weaknesses
\citep{siek2015blame,siek2010threesomes,garcia2013calculating,kuhlenschmidt2018efficient,siek2012interpretations,garcia2014deriving}. This
paper presents a new cast representation, named \emph{hypercoercions},
that is good for both mechanized metatheory and good for
implementation. Recall that it is the composition operator in these
systems that enables space-efficiency by compressing casts.  The
composition operator for hypercoercions is structurally recursive and
hypercoercions are suggestive of a bit-level representation that
minimizes the need for pointers and fits all first-order casts into 64
bits.  Of course, an alternative cast representation must not change
the result of programs. This paper presents steps toward a framework
for proving the equivalence of cast calculi and, in particular, proves
that an abstract machine using hypercoercions is equivalent to an
abstract machine using standard casts.

%
%For example, translation to intermediate languages can wrap some expressions 
%with casts, including expressions at tail positions. Thus, mutually 
%tail-recursive functions in source language can consume unbounded space at 
%runtime. There are more examples of space leaking in \citet{herman2010space}.

The current state of the art regarding cast representations includes
\begin{itemize}
\item threesomes \citep{siek2010threesomes},
\item supercoercions \citep{garcia2014deriving}, and
\item coercions in normal form
  \citep{siek2012interpretations,siek2015blame}.
\end{itemize}
Threesomes and supercoercions are good for mechanized metatheory
because their compose operators are structurally recursive, making
them easy to define in a proof assistant such as Agda. In contrast,
the coercions in normal form have compose operators that are not
structurally recursive, which makes it more difficult to define in
Agda, requiring what amounts to an explicit proof of termination.
%
On the other hand, coercions in normal form are easier to understand
than threesomes (with a strange labelled bottom type), and
supercoercions (with 10 different kinds).

%% There are two best studied ways to represent cast space-efficiently:
%% coercions
%% \citep{herman2010space}\citep{siek2009exploring}\citep{siek2012interpretations}
%% \citep{siek2015blame} and threesomes
%% \citep{siek2010threesomes}\citep{garcia2013calculating}. They have
%% different strength and weakness. \citet{garcia2013calculating} and
%% \citet{siek2015blame} claim that coercion is easy to understand, but
%% hard to implement, while threesome is easy to implement, but hard to
%% understand. \citet{garcia2013calculating} in addition claims that
%% coercion is more difficult to reason about formally because its
%% compose function is not structurally recurisive. For instance, the
%% developers of Grift \citep{kuhlenschmidt2018efficient} report that it
%% is tricky to convince their proof assistants that the compose function
%% terminates.

We present a new cast representation, hypercoercion, which has a structurally 
recursive composition. When we prove theorems about hypercoercion in Agda, the 
definition ``just works''. 
Besides, We believe hypercoercion is about as easy to understand as coercion 
because their syntax definitions are similar. For instance, the 
coercion approach represents $ 
\cOOcast{\TOOdyn}{l}{\cOOcast{\POOunit}{l'}{\TOOdyn}} $ as $ \POOunit 
?^{l} ; \POOunit\! $, while the hypercoercion approach gives $ 
\hyperCoercionC{?^l}{\POOunit}{!} $. In Section~\ref{sec:background}, we review 
coercion together with cast calculus 
\citep{siek2009exploring}\citep{kuhlenschmidt2018efficient}. And we 
present hypercoerion in Section~\ref{sec:hypercoercion-definition}.

Another strength of hypercoercion over coercion is that hypercoercion might 
have a more compact memory representations. In Grift 
\citep{kuhlenschmidt2018efficient}, the longest coercions are represented with
two cons boxes. But for hypercoercion, we are optimistic that one cons box 
suffices. Using less cons boxes saves memory as well as running time.
\todo[inline]{
	Could Jeremy help explain explaining why hypercoercion is more compact in 
	memory?
	Cons boxes seem to be non-standard, but I don't know a better name.
 -KC}

Many of the existing research on coercions or threesomes include proof(s) of 
correctness, that is, the evaluation (partial) function of cast calculus $ eval 
$ is equal to their space-efficient evaluation ($ eval_{*} $).
Despite the striking similarity of these proofs, they are all built from ground 
up.

We present a framework for proving correctness of cast reprensentation 
(Section~\ref{sec:framework}), and use it to prove the correctness of 
hypercoercion (Section~\ref{sec:hypercoercion-correctness}). We are optimistic 
that this framework is general enough to include coercion and threesome.

%Some readers might know that there are four blame strategies: $ \{Lazy, 
%Eager\} 
%\times \{D, UD\} $ \citep{siek2009exploring}. Eager strategies detect more 
%potential type errors, but 
%blame more programs. $ D $ and $ UD $ assign blame labels differently.

We focus on lazy cast strategies because we suspect that they are more
efficient that eager strategies, which is why lazy was chosen in the
Grift compiler \citep{Kuhlenschmidt:2019aa}, and because
\citet{new2019gradual} show that eager strategies are incompatible
with $\eta$-equivalence of functions. We are interested in the D blame
tracking strategy because it has a more straightforward notion safe
cast compared to UD \citep{siek2009exploring}, which is why D was
chosen Grift compiler. However, we are also intersted in UD because it
plays a prominent role in the gradual typing literature
\citep{wadler2009well}.

The primary contributions of this paper are:
\begin{itemize}
\item hypercoercions, a new space-efficient cast representation, which
  has a structurally recursive composition and a more compact memory
  representation.
\item a framework in Agda for proving correctness of cast
  represantations.  We conjecture that both coercion and
  threesome can fit into this framework.
\item a formal proof that hypercoercion respects the semantics
  of the \lazyD{} cast calculus.
\end{itemize}

%Around 2006, many work on integrating dynamic typing and static typing emerge, 
%notably gradual types \citep{siek2006gradual} and hybrid types 
%\citep{flanagan2006hybrid}. 
%Later \citet{wadler2009well} introduces Blame Calculus, an intermediate 
%language for gradual types and hybrid types. 
%Blame Calculus throws a blame label when a runtime type-check fails.
%Labels provide useful debugging information, therefore are necessary for 
%practical use of gradually typed languages.
%
%Implementing gradually typed languages naively on top of blame calculus, 
%however, suffices sever space leak.
%For example, translation to Blame Calculus can wrap some expressions with code 
%that performs runtime type-checking (casts), including expressions at 
%tail positions. Thus, mutually tail-recursive functions in source language can 
%consume unbounded space at runtime. There are more examples of space leaking 
%in 
%\citet{herman2010space}.
%
%In 2007, \citet{herman2010space} proposes a solution to this problem by 
%representing casts with coercion \citep{henglein1994dynamic}, 
%which can be composed and normalized. Every coercion looks like a list of 
%trees. For normal coercion, the length of the list is bounded by a small 
%constant. And the depths of trees do not grow during composition and
%normalization. These two facts implies that coercion is a space-efficient cast 
%representation. This coercion, however, does not support blame 
%tracking. 
%
%After that, many efforts have been made to combine blame tracking and 
%space efficiency. 
%
%\citet{siek2009exploring} present another intermediate langauge for gradually 
%typed language, cast calculus \castCalculus\ and achieve both blame-tracking 
%and space-efficiency for this language by decorating \citet{herman2010space}'s 
%coercions with labels. \castCalculus\ is simply typed lambda calculus extended 
%with the dynamic type and cast expressions. The labels in \castCalculus{}, 
%unlike their counterparts in the Blame Calculus, have no polarity. 
%They also propose that there are four blame strategies for \castCalculus{}:
%$ \{Lazy, Eager\} \times \{D, UD\} $. Eager strategies detect more potential 
%type errors, but blame more programs. $ D $ and $ UD $ assign blame labels 
%differently. They claim that their \lazyUD\ cast calculus is the same as the 
%Blame Calculus, modulo label polarity, and prove that their their \lazyUD\ 
%coercion calculus simulates their \lazyUD{} cast calculus. It is unknown, 
%however, whether the simulation in the other direction also works.
%\todo[inline]{Is it important to this paper whether there is a
%simulation in the other direction? -JS
%\\ --- \\
%I guessed to say coercion is correct the other direction is 
%necessary. -KC}
%
%\citet{siek2010threesomes} proposes another approach to combine blame 
%tracking and space efficiency. Their solution represents casts with 
%threesomes, 
%a novel cast representation. Threesome is proved space-efficient as well.
%They prove that their calculus bisimulates the \lazyUD{} coercion 
%calculus in \citet{siek2009exploring} and the Blame Calculus in 
%\citet{wadler2009well}.
%
%\citet{siek2012interpretations} present a revised coercion calculi.
%They simplify the coercion calculi by only 
%working with normal coercions (i.e. coercions in normal form). 
%They conjecture that their revised coercion calculi bisimulate the 
%corresponding cast calculi.
%
%\citet{garcia2013calculating} develops
%threesome for other blame strategies. After studying in depth the connection 
%between coercion and threesome,
%he claims that normal coercion with labels is easy to understand but hard to 
%implement and hard to reason about, and that 
%threesome with labels, however, is easy to implement and easy to reason about 
%but hard to understand. 
%His claim is mostly affirmed by the group of people who develop threesome 
%(\citet{siek2015blame}).
%His threesome calculi are equivalent to the coercion calculi of
%\citet{siek2009exploring} because the former ones are derived from the latter 
%ones.
%
%\citet{siek2015blame} revisit the coercion-based approach. They revise
%the \lazyUD{} coercion in \citet{siek2012interpretations} and prove
%that their revised coercion calculus bisimulates the \lazyUD{} Blame
%Calculus \citet{wadler2009well}.
%
%Last year, \citet{kuhlenschmidt2018efficient} present Grift, a compiler for a 
%gradually typed language of the same name. 
%This implementation is based on the \lazyD{} coercion in 
%\citet{siek2012interpretations}.
%Their result suggests that normal coercion is easy enough to implement in a 
%compiler. 
%
%So far, it seems that normal coercion is the best way to combine 
%space-efficiency and blame-tracking: normal coercion is claimed easy to 
%understand \citep{garcia2013calculating}\citep{siek2015blame} and is shown 
%easy enough to implement in a compiler \citep{kuhlenschmidt2018efficient}.
%Normal coercion is still not satisfactory, however, in at least two aspects.
%Firstly, as we mentioned above, it is hard to reason about normal coercion. 
%The 
%major difficulty is from its non-structurally recursive composition.
%For instance, the developers of Grift report it is tricky to convince
%their proof assistants that the composition terminates.
%
%Secondly, the definition of normal coercion lies unnecessarily on top 
%of coercion: all coercions are normal when they are initially constructed, and 
%\citet{siek2012interpretations} have shown an function that 
%composes and produces normal coercions.
%%
%\todo[inline]{Many researchers would view this as a positive thing,
%  not a negative. For example, the notion of normal forms comes up
%  over and over again in mathematics and logic. -JS
%\\ --- \\
%I guess having a notion of normal forms is more convenient when a 
%notion of normalization is convenient. But we don't even need a notion of 
%normalization for cast representations. -KC
%\\ --- \\
%FWIW, I agree with you. But the point is that the argument is not
%likely to be convincing to the reader, so we're probably better off
%not wasting space on it. -JS
%}
%
%Perhaps unsurprisingly, threesome has a structurally recursive compose and a 
%self-standing definition. Together with \citet{garcia2013calculating}'s 
%work, they suggest that there might be a cast representation whose definition 
%is self-standing, and whose composition is structurally recursive. 
%supercoercion introduced in \citet{garcia2013calculating} could be a promising 
%candidate. It is believed even more complicated than threesome 
%\citep{siek2015blame}. In fact, it uses 10 constructors to deal with an 
%elementary type system with only base types and 
%function types. And four constructors are directly related to function types. 
%Thus, supercoercion might not scale very well to more sophisticated type 
%systems. 
%
%We present hypercoercion, a cast representation whose composition is 
%structurally recursive, and whose definition is self-standing. What's more, it 
%should be at 
%least as easy to implement and understand as normal coercions, because there 
%is 
%a clear connection between them, as we will show in Section 
%\ref{sec:hypercoercion}.
%Our hypercoercion considers sum types and product types, which are not 
%considered in all proofs above. Adding each of them requires us to add only 
%one 
%new constructor to a component of hypercoercion. This suggests that 
%hypercoercion might scale better than supercoercion.
%
%We prove formally that the \lazyD{} (resp. \lazyUD{}) hypercoercion calculus 
%bisimulates the \lazyD{} (resp. \lazyUD{}) cast calculus. This is 
%the first bisimulation proof for the \lazyD{} cast calculus as far as we know.
%We care more about lazy strategies because recently, \citet{new2019gradual} 
%show that \eager{} strategies are incompatible with $\eta$-equivalence of 
%functions, which suggest that these strategies are not very ideal. 
%\todo[inline]{Explain why we care about \lazyD{} and \lazyUD{}. -JS
%  \\ --- \\ Explained, not sure if it suffice. -KC
%  \\ --- \\
%  That's good that you explained why lazy instead of eager.
%  (Also, you should add that we are interested in lazy because
%  that's what Grift does.)\\
%  But you haven't explained why we are interested in both D and UD. -JS
%}
%
%The structure of this paper is as follows. 
%Section \ref{sec:blame-calculus} reviews the state-of-art of $Lazy$ cast 
%calculi. 
%Section \ref{sec:coercion-calculus} reviews the state-of-art of $Lazy$ 
%Coercion 
%Calculi. 
%In section \ref{sec:hypercoercion} we present hypercoercion.
%Section \ref{sec:conclude} concludes.
%\todo[inline]{Lots of missing sections. -JS}

@@

In this section, we review lazy cast calculi (\ref{sec:cast-calculi}) and
normal coercion (\ref{sec:coercion-calculus}).

\subsection{Cast Calculi}
\label{sec:cast-calculi}
\begin{figure}
	Syntax
	\[
	\begin{array}{rclr}
	\stxrule{T}{types}{
		\star \mid{}
		P
	}
	\stxrule{P}{pre-types}{
		\POOunit \mid
		\POOfun{T_1}{T_2} \mid
		\POOprod{T_1}{T_2} \mid
		\POOsum{T_1}{T_2}
	}
	\stxrule{e}{terms}{
		\eOOvar{x} \mid{}
		\eOOsole{} \mid{}
		\eOOlam{T_1}{T_2}{x}{e} \mid
		\eOOapp{e_1}{e_2} \mid
		\eOOcons{e_1}{e_2} \mid
		\eOOcar{e} \mid
		\eOOcdr{e}
	}
	\stxrulecont{
		\eOOinl{e} \mid
		\eOOinr{e} \mid
		\eOOcase{e_1}{e_2}{e_3} \mid
		\eOOcast{e}{T_1}{l}{T_2} \mid
		\eOOblame{l}
	}
	\stxrule{v}{values}{
		\vOOtt{} \mid
		\vOOfun{\rho}{x}{e} \mid
		\vOOcons{v_1}{v_2} \mid
		\vOOinl{v} \mid
		\vOOinr{v} \mid		
		\vOOcast{v}{c}
	}
	\stxrule{c}{casts}{
		\cOOcast{T_1}{l}{T_2}
	}
	\stxrule{I}{injectable types (\lazyD)}{
		P
	}
	\stxrule{I}{injectable types (\lazyUD)}{
		\POOunit \mid
		\POOfun{\star}{\star} \mid
		\star \times \star \mid
		\star + \star
	}
	\stxrule{o}{observations}{
		\oOOinj \mid
		\oOOsole \mid
		\oOOfun \mid
		\oOOcons \mid
		\oOOinl \mid
		\oOOinr \mid
		\oOOblame{l}
	}
	\end{array}
	\]

	Consistency
	\fbox{$ T_1 \sim T_2 $}
	\begin{gather*}
	\inference{}{
		\star \sim \star
	} \quad
	\inference{}{
		\star \sim P
	} \quad
	\inference{}{
		P \sim \star
	} \\
	\inference{}{
		\iota \sim \iota
	} \quad
	\inference{
		S_1 \sim S_2 &
		T_1 \sim T_2
	}{
		S_1 \rightarrow T_1 \sim S_2 \rightarrow T_2
	} \quad
	\inference{
		S_1 \sim S_2 &
		T_1 \sim T_2
	}{
		S_1 \times T_1 \sim S_2 \times T_2
	} \quad
	\inference{
		S_1 \sim S_2 &
		T_1 \sim T_2
	}{
		S_1 \plus T_1 \sim S_2 \plus T_2
	}
	\end{gather*}

@@

	Term typing
	\fbox{$ \judgetype{\Gamma}{e}{T} $}
	\begin{gather*}
		\inference{
			\Gamma \vdash e : T_1 & T_1 \sim T_2
		}{
			\judgetype{\Gamma}{\eOOcast{e}{T_1}{l}{T_2}}{T_2}
		} \quad
		\inference{
		}{
			\judgetype{\Gamma}{\eOOblame{l}}{T}
		}
	\end{gather*}

	Value typing \fbox{$ v : T $}
	\begin{gather*}
	\dots \quad
	\inference{
		v : I
	}{
		\vOOcast{v}{\cOOcast{I}{l}{\TOOdyn}} : \TOOdyn
	}
	\quad
	\inference{
		v : P_1 &
		P_1 \smile P_2
	}{
		\vOOcast{v}{\cOOcast{P_1}{l}{P_2}} : P_2
	}
	\end{gather*}

\figref{fig:blame-static} defines the syntax of cast calculi and its
static semantics. We extend the syntax of \citet{siek2009exploring}
with sum and product types.  Let $ T $ range over types. A type is
either the dynamic type $ \star $ (a.k.a. $ \mathtt{Dyn} $, $
\mathbb{?} $, or $ \mathtt{Unknown} $), or a type with a type
constructor at the top, that is, a pre-type, ranged over by $ P $. We
include just one base type, $\POOunit$. The other pre-types are
functions, products, and sums.

As usual we write $ T_1 \sim T_2 $ when $ T_1 $ and $ T_2 $ are
consistent. The intuition of $ T_1 \sim T_2 $ is that $ T_1 $ and $
T_2 $ have no conflicting type information. Two types are consistent
if one of them is $ \star $, or they have the same top-most type
constructor and the corresponding sub-parts are
consistent. Consistency is reflexive and symmetric, but not
transitive.

We write $ T_1 \smile T_2 $ when $ T_1 $ and $ T_2 $ are
shallowly-consistent, that is, if one of them is $ \star $, or they
have the same type constructor at the top. Shallow-inconsistency is
the root of all blame in lazy cast strategies -- casting a value to a
shallowly inconsistent type leads to a blame. Shallow-consistency is
reflexive, symmetric, but not transitive.

Let $ e $ range over terms. Most terms are standard for a simply-typed
lambda calculus. We annotate the codomain of a lambda abstraction
explicitly because we refer to it in the dynamic semantics. Becuase
this is a cast calculus, we have two additional terms: cast
expressions and blame expressions. Cast expressions perform runtime
type checks and blame expressions raise an error.

@@

	Syntax
	\[
	\begin{array}{rclr}
	\stxrule{r}{cast results}{
		\rOOsucc{v} \mid
		\rOOfail{l}
	}
	\stxrule{s}{states}{
		\sOOinspect{e}{\rho}{\kappa} \mid{}
		\sOOreturn{v}{\kappa} \mid{}
		\sOOhalt{o}
	}

>>>>

\newcommand{\cOOcast}[3]{#1 \Rightarrow^{#2} #3}

@@

\newcommand{\hiredrule}[3]{\highlight{#1} & \highlight{\longmapsto_\mathcal{C}} & \highlight{#2} & \highlight{#3} \\}

@@

\newcommand{\highlight}[1]{{\color{red} #1}}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Designing a space-efficient cast representation that is good for
  both mechanized metatheory and implementation is
  challenging. Existing solutions are good for one or the other. This
  paper presents a new cast representation, named hypercoercions, that
  is good for both. On the way to proving the correctness of
  hypercoercions, this paper also makes progress on a general
  framework for proving the correctness of cast representations.
\end{abstract}

@@

Around 2006, several groups of researchers proposed ways to integrate
dynamic typing and static typing, notably gradual typing
\citep{siek2006gradual}, hybrid typing \citep{flanagan2006hybrid},
migratory typing \citep{Tobin-Hochstadt:2006fk}, and multi-language
iteroperability \citep{Gray:2005ij,Matthews:2007zr}. Researchers
usually define the semantics of gradually typed languages by
translation to an intermediate language with casts, such as the blame
calculus \citep{wadler2009well} and other cast calculi
\citep{siek2009exploring}. Unfortunately, straightforward
implementations of casts on higher-order values (functions, objects,
etc.) impose significant runtime overheads that can change the
asymptotic time and space complexity of a program
\citep{herman2010space}. There are several known space-efficient cast
representations, with various strengths and weaknesses
\citep{siek2015blame,siek2010threesomes,garcia2013calculating,kuhlenschmidt2018efficient,siek2012interpretations,garcia2014deriving}.
The current state of the art includes
\begin{itemize}
\item threesomes \citep{siek2010threesomes},
\item supercoercions \citep{garcia2014deriving}, and
\item coercions in normal form
  \citep{siek2012interpretations,siek2015blame}.
\end{itemize}
Recall that in these systems, casts are compressed using a composition
operator.  Threesomes and supercoercions are good for mechanized
metatheory because their compose operators are structurally recursive,
making them easy to define in a proof assistant such as Agda. In
contrast, the coercions in normal form have compose operators that are
not structurally recursive, which makes it more difficult to define in
Agda, requiring what amounts to an explicit proof of termination.
%
On the other hand, coercions in normal form are easier to understand
than threesomes (with a strange labelled bottom type), and
supercoercions (10 different kinds).

This paper presents a new cast representation, named
\emph{hypercoercions}, that is good for both mechanized metatheory and
good for implementation. The composition operator for hypercoercions
is defined by structural recursion and hypercoercions are suggestive
of a bit-level representation that minimizes the need for pointers and
fits all first-order casts into 64 bits. We present two flavors of
hypercoercions to support the two blame tracking strategies from the
literature: D and UD~\citep{siek2009exploring}. We are interested in
the D blame tracking strategy because it comes with a more
straightforward notion safe cast compared to
UD~\citep{siek2009exploring}, which is why D was chosen Grift
compiler. We are also intersted in UD because it plays a prominent
role in the gradual typing literature \citep{wadler2009well}. The
semantics of casts can also be lazy or
eager~\citep{siek2009exploring}. In this paper we focus on lazy cast
strategies because we suspect that they are more efficient than eager
strategies and because \citet{new2019gradual} show that eager
strategies are incompatible with $\eta$-equivalence of functions.

Of course, an alternative cast representation must be proved
correct. This paper presents steps toward a general framework for
proving equivalence of cast calculi and, in particular, proves that an
abstract machine using \lazyD{} hypercoercions is equivalent to an
abstract machine using standard \lazyD{}
casts~\citep{siek2009exploring}. We conjecture that the framework can
be generalized to \lazyUD{} and that it can be applied to coercions in
normal form and threesomes.

To summarize, the primary contributions of this paper are:
\begin{itemize}
\item hypercoercions, a new space-efficient cast representation, which
  has a structurally recursive composition and a more compact memory
  representation.
\item a framework in Agda for proving the correctness of cast
  represantations.
\item a formal proof that hypercoercion respects the semantics
  of the \lazyD{} cast calculus.
\end{itemize}

In Section~\ref{sec:background} we review cast calculi and coercions.
We present hypercoercions in
Section~\ref{sec:hypercoercion-definition}.  We present a framework
for proving correctness of cast reprensentations in
Section~\ref{sec:framework} and use it to prove the correctness of
\lazyD{} hypercoercions in
Section~\ref{sec:hypercoercion-correctness}.

@@

In this section, we review lazy cast calculi
(Section~\ref{sec:cast-calculi}) and coercion in normal form
(Section~\ref{sec:coercion-calculus}).

\subsection{Cast Calculi}
\label{sec:cast-calculi}

\begin{figure}
	Syntax
	\[
	\begin{array}{rclr}
	\stxrule{T}{types}{
		\star \mid{}
		P
	}
	\stxrule{P}{pre-types}{
		\POOunit \mid
		\POOfun{T_1}{T_2} \mid
		\POOprod{T_1}{T_2} \mid
		\POOsum{T_1}{T_2}
	}
	\stxrule{e}{terms}{
		\eOOvar{x} \mid{}
		\eOOsole{} \mid{}
		\eOOlam{T_1}{T_2}{x}{e} \mid
		\eOOapp{e_1}{e_2} \mid
		\eOOcons{e_1}{e_2} \mid
		\eOOcar{e} \mid
		\eOOcdr{e}
	}
	\stxrulecont{
		\eOOinl{e} \mid
		\eOOinr{e} \mid
		\eOOcase{e_1}{e_2}{e_3} \mid
		\eOOcast{e}{T_1}{l}{T_2} \mid
		\eOOblame{l}
	}
	\end{array}
	\]

	Consistency
	\fbox{$ T_1 \sim T_2 $}
	\begin{gather*}
	\inference{}{
		\star \sim \star
	} \quad
	\inference{}{
		\star \sim P
	} \quad
	\inference{}{
		P \sim \star
	} \\[1ex]
	\inference{}{
		\iota \sim \iota
	} \quad
	\inference{
		S_1 \sim S_2 &
		T_1 \sim T_2
	}{
		S_1 \rightarrow T_1 \sim S_2 \rightarrow T_2
	} \quad
	\inference{
		S_1 \sim S_2 &
		T_1 \sim T_2
	}{
		S_1 \times T_1 \sim S_2 \times T_2
	} \quad
	\inference{
		S_1 \sim S_2 &
		T_1 \sim T_2
	}{
		S_1 \plus T_1 \sim S_2 \plus T_2
	}
	\end{gather*}

	Term typing
	\fbox{$ \judgetype{\Gamma}{e}{T} $}
	\begin{gather*}
          \dots \qquad
		\inference{
			\Gamma \vdash e : T_1 & T_1 \sim T_2
		}{
			\judgetype{\Gamma}{\eOOcast{e}{T_1}{l}{T_2}}{T_2}
		} \quad
		\inference{
		}{
			\judgetype{\Gamma}{\eOOblame{l}}{T}
		}
	\end{gather*}

\paragraph{Syntax and Static Semantics}

The syntax and static semantics is the same for the \lazyD{} and
\lazyUD{} cast calculi, and is reviewed in \figref{fig:blame-static}.
As usual, the important features are the cast expressions,
$\eOOcast{e}{T_1}{l}{T_2}$, which are responsible for runtime type
checking, and blame expressions, $\eOOblame{l}$, that raise errors.
The syntax and static semantics is the same that of
\citet{siek2009exploring} except for a few minor exceptions:
\begin{itemize}
\item We add sum, product, and unit types.
\item We separate types into those with a type constructor at the top,
  the \emph{pretypes} ($\POOunit$, functions, products, and sums),
  versus the dynamic type $\star$ (a.k.a. $ \mathtt{Dyn} $ or
  $\mathbb{?}$).
\item We annotate the codomain of a lambda abstraction explicitly
  because we refer to it in the dynamic semantics. 
\end{itemize}
As usual, the source $T_1$ and target types $T_2$ of a cast
$\eOOcast{e}{T_1}{l}{T_2}$ are required to be consistent, written $T_1
\sim T_2$. The consistency relation is standard and defined in
\figref{fig:blame-static}.

%% defines the syntax of the cast calculus and
%% its static semantics. We extend the syntax of
%%  with sum and product types.  Let $ T $ range
%% over types. A type is either the dynamic type $ \star $ (a.k.a. $
%% \mathtt{Dyn} $, $ \mathbb{?} $, or $ \mathtt{Unknown} $), or a type

%% As usual we write $ T_1 \sim T_2 $ when $ T_1 $ and $ T_2 $ are
%% consistent. The intuition of $ T_1 \sim T_2 $ is that $ T_1 $ and $
%% T_2 $ have no conflicting type information. Two types are consistent
%% if one of them is $ \star $, or they have the same top-most type
%% constructor and the corresponding sub-parts are
%% consistent. Consistency is reflexive and symmetric, but not
%% transitive.

%% We write $ T_1 \smile T_2 $ when $ T_1 $ and $ T_2 $ are
%% shallowly-consistent, that is, if one of them is $ \star $, or they
%% have the same type constructor at the top. Shallow-inconsistency is
%% the root of all blame in lazy cast strategies -- casting a value to a
%% shallowly inconsistent type leads to a blame. Shallow-consistency is
%% reflexive, symmetric, but not transitive.

%% Let $ e $ range over terms. Most terms are standard for a simply-typed
%% lambda calculus. We annotate the codomain of a lambda abstraction
%% explicitly because we refer to it in the dynamic semantics. Becuase
%% this is a cast calculus, we have two additional terms: cast
%% expressions and blame expressions. Cast expressions perform runtime
%% type checks and blame expressions raise an error.

\paragraph{Dynamic Semantics}

The dynamic semantics of a cast calculus is typically defined with a
reduction semantics. Here we use a CEK machine
\citep{felleisen1986control} instead because the first author is more
familiar with CEK machines and believes that a CEK machine is more
convenient to use for the space-efficient semantics in
Section~\ref{sec:framework:cek} (space-efficiency is about compressing
continuations). So using an abstract machine for the cast calculi in
this section makes it easier prove correctness of the space-efficient
machines. Of course, one should prove that the abstract machine
presented here is equivalent to the standard reduction semantics for
cast calculi, but we have not yet done so.

Fig.~\ref{fig:machine-cekc} defines the transition relation of the CEK
machine and Fig.~\ref{fig:state} gives a grammar for machine states
$s$, including a definition of values and value typing. The
transitions involving casts are highlighted in red and described in
more detail below. The other transitions are standard for a CEK
machine for an extended simply typed lambda calculus.
%
Recall that CEK machine involves two kinds of transitions, (1) those
that dive further into an expression (looking for a redex) and push an
entry onto the continuation, and (2) those that return a value to the
current continuation and possibly perform a computation.
Corresponding to (1) and (2), the machine state is either in an
$\mathtt{Eval}$ or $\mathtt{Cont}$ configuration, respectively.
Additionally, there is the $\mathtt{Halt}$ configuration.

The transition relation $\judgeCreduce{s}{s}$ is parameterized over
$applyCast$ to allow for the differences between D and UD.
%
When evaluating a cast expression, the machine moves the cast to the
continuation and evaluates the inner expression.
%% Other
%% transition rules starting from $ Eval $ states are standard.
To apply a casted function, the machine first casts $ v_1 $, the
operand, then applies the casted operand to $ v_2 $, the underlying
function, and finally cast the result of the function application.
%
To take out the first (resp. second) part of a casted pair, the
machine firstly take out the first (resp. second) part of $ v $, the
underlying pair, and cast the result.
%
In case splitting, if the target value is a left injection, the
machine moves to a state that will apply the first continuation
function to the value inside the left injection. The case for right
injection is similar.
%
When the target value is a casted sum, the machine moves the cast from
the target value to continuations functions.
%
To cast a value, the machine invokes $applyCast$ on the value.  If the
cast succeeds, the machine returns the result to the next
continuation.  If the cast fails, the machine halts with the blame
label.
%
%% When the continuation is $ stop $, the machine enters a halting
%% state. The When the machine is in a halting state, it stays in the
%% same state.

The reflexive transitive closure of reduction ($
\judgeCreduceTrans{s}{s} $) and evaluation ($ eval_\mathcal{C}(e) $)
are standard~\citep{felleisen03:_pllc}.

\begin{figure}
  Machine state and other runtime data structures
  \[
  \begin{array}{rclr}
	\stxrule{v}{values}{
		\vOOtt{} \mid
		\vOOfun{\rho}{x}{e} \mid
		\vOOcons{v_1}{v_2} \mid
		\vOOinl{v} \mid
		\vOOinr{v} \mid		
		\vOOcast{v}{c}
	}
	\stxrule{c}{casts}{
		\cOOcast{T_1}{l}{T_2}
	}
	\stxrule{I}{injectable types (\lazyD)}{
		P
	}
	\stxrule{I}{injectable types (\lazyUD)}{
		\POOunit \mid
		\POOfun{\star}{\star} \mid
		\star \times \star \mid
		\star + \star
	}
	\stxrule{o}{observations}{
		\oOOinj \mid
		\oOOsole \mid
		\oOOfun \mid
		\oOOcons \mid
		\oOOinl \mid
		\oOOinr \mid
		\oOOblame{l}
	}
	\stxrule{r}{cast results}{
		\rOOsucc{v} \mid
		\rOOfail{l}
	}
	\stxrule{s}{states}{
		\sOOinspect{e}{\rho}{\kappa} \mid{}
		\sOOreturn{v}{\kappa} \mid{}
		\sOOhalt{o}
	}
	\stxrule{\kappa}{continuations}{
		\kOOmt \mid
		\kOOconsI{e}{\rho}{\kappa} \mid
		\kOOconsII{v}{\kappa} \mid
		\kOOinl{\kappa} \mid
		\kOOinr{\kappa}
	}
	\stxrulecont{
		\kOOappI{e}{\rho}{\kappa} \mid
		\kOOappII{v}{\kappa} \mid
		\kOOcar{\kappa} \mid
		\kOOcdr{\kappa} \mid
		\kOOcaseI{e_1}{e_2}{\rho}{\kappa}
	}
	\stxrulecont{	
		\kOOcaseII{v}{e}{\rho}{\kappa} \mid
		\kOOcaseIII{v_1}{v_2}{\kappa} \mid
		\kOOcast{c}{\kappa}
	}
  \end{array}
  \]

@@

	Value typing \fbox{$ v : T $}
	\begin{gather*}
	\dots \qquad
	\inference{
		v : I
	}{
		\vOOcast{v}{\cOOcast{I}{l}{\TOOdyn}} : \TOOdyn
	}
	\quad
	\inference{
		v : P_1 &
		P_1 \smile P_2
	}{
		\vOOcast{v}{\cOOcast{P_1}{l}{P_2}} : P_2
	}
	\end{gather*}
        \caption{Definition of machine state and auxilliary data
          structures.}
        \label{fig:state}
\end{figure}

@@

	\[
	\begin{array}{rclr}
	\end{array}
	\]

	Reduction \fbox{$ \judgeCreduce{s}{s} $}
	\[
	\begin{array}{rclr}
		\hiredrule{
		\sOOinspect{\eOOcast{e}{T_1}{l}{T_2}}{\rho}{\kappa}
	}{
		\sOOinspect{e}{\rho}{\langle\cOOcast{T_1}{l}{T_2}\rangle\kappa}
	}{}
